Finch: Sparse and Structured Array Programming with Control Flow

### 2024-04-03 sections/eval.tex
Update on Overleaf.

There are three main approaches to SpGEMM \cite[Section 2.2]{zhang2021gamma}.
%
The inner-products algorithm takes dot products of corresponding rows and columns, while the outer-products algorithm sums the outer products of corresponding columns and rows.
%
Gustavson's algorithm sums the rows of $B$ scaled by the corresponding nonzero columns in each row of $A$.
%
Inner products is known to be asymptotically less efficient than the others, as we must do a merge operation to compute each of the $O(MN)$ entries in the output \cite{ahrens2022autoscheduling}.

Figure~\ref{fig:spgemm_listing} implements all three approaches in Finch, and Figure~\ref{fig:spgemm} compares the performance of Finch to TACO.
%
Note that these algorithms mainly differ in their loop order, but that different datastructures can be used to support the various access patterns induced.
%
Although a sparse bytemap has a dense memory footprint, we use it in our Finch implemenation of Gustavson's for the smaller $O(M)$ intermediate.
%
In our Finch implementation of outer products, we use a sparse hash table, as it is fully-sparse and randomly accessible.
%
Unfortunately, TACO does not support multidimensional sparse workspaces, only supporting outer-products with a dense output, a limitation with asymptotic consequences.
%
It's also worth mentioning that the bytemap format in TACO's Gustavson's implementation is hard-wired, whereas Finch's programming model allows us to write algorithms with explicit temporary formats and transpositions.

As depicted in Figure~\ref{fig:spgemm}, Finch achieves comparable performance with TACO on smaller matrices when we use the same datastructures, and significant improvements when we use better datastructures. Finch outperforms TACO on larger matrices, with an average speedup of 1.05.

@@

\begin{figure}
	\includegraphics[width=0.5\linewidth]{spgemm_small_speedup_log_scale.png}%
	\includegraphics[width=0.5\linewidth]{spgemm_joel_speedup.png}
    \vspace{-8pt}
    \caption{A comparison of several matrix multiplication algorithms between
    Finch and Taco. On left, we use smaller matrices, ordered from small to big
    dimension. Note that inner products necessarily requires $O(n^2)$ work and
    TACO's outer products format is dense. Finch can use a sparse outer products
    format and thus has an asymptotic advantage that becomes evident as the
    output dimensions grow. On right, we use only gustavson's algorithm and
    compare on larger matrices.}
    \label{fig:spgemm}
    \vspace{-12pt}
\end{figure}

@@

We used finch to implement two fundamental graph algorithms: Breadth-First Search (BFS) and Bellman-Ford single-source shortest path. Our BFS implementation and graphs datasets are taken from Yang et al.~\cite{yang_implementing_2018}, including road networks, characterized by bounded node degrees and long diameters, and scale-free graphs, where node degree distribution follows a power-law distribution and diameters are short.

@@

Figure \ref{fig:graph_result} compares performance to Graphs.jl, an open-source Julia library, and the LAGraph Library, which implements graph algorithms in the language of linear algebra using GraphBLAS\cite{mattson2019lagraph}. For the BFS algorithm, direction-optimization notably enhances performance for scale-free graphs. Although GraphBLAS uses hardwired optimizations, Finch is competitive on these benchmarks. On Bellman-Ford (with path lengths and shortest-path tree), Finch's support for multiple outputs, sparse inputs, and masks leads to superior performance over GraphBLAS (average speedup of 1.05). Note that we did not include GAP-road as it took too long to run.

>>>>

There are three main approaches to SpGEMM \cite[Section 2.2]{zhang2021gamma}.
%
The inner-products algorithm takes dot products of corresponding rows and columns, while the outer-products algorithm sums the outer products of corresponding columns and rows.
%
Gustavson's algorithm sums the rows of $B$ scaled by the corresponding nonzero columns in each row of $A$.
%
Inner-products is known to be asymptotically less efficient than the others, as we must do a merge operation to compute each of the $O(MN)$ entries in the output \cite{ahrens2022autoscheduling}.

Figure~\ref{fig:spgemm_listing} implements all three approaches in Finch, and Figure~\ref{fig:spgemm} compares the performance of Finch to TACO.
%
Note that these algorithms mainly differ in their loop order, but that different datastructures can be used to support the various access patterns induced.
%
Although a sparse bytemap has a dense memory footprint, we use it in our Finch implemenation of Gustavson's for the smaller $O(M)$ intermediate.
%
In our Finch implementation of outer products, we use a sparse hash table, as it is fully-sparse and randomly accessible.
%
Unfortunately, TACO does not support multidimensional sparse workspaces, only supporting outer-products with a dense output, a limitation with asymptotic consequences.
%
It is also worth mentioning that the bytemap format in TACO's Gustavson's implementation is hard-wired, whereas Finch's programming model allows us to write algorithms with explicit temporary formats and transpositions.

As depicted in Figure~\ref{fig:spgemm}, Finch achieves comparable performance with TACO on smaller matrices when we use the same datastructures, and significant improvements when we use better datastructures. Finch outperforms TACO on larger matrices, with an average speedup of 1.25. %arith-mean

@@

\begin{figure}
	\includegraphics[width=0.5\linewidth]{spgemm_small_speedup_log_scale.png}%
	\includegraphics[width=0.5\linewidth]{spgemm_joel_speedup.png}
    \vspace{-8pt}
    \caption{A comparison of several matrix multiplication algorithms between
    Finch and Taco. On left, we use smaller matrices, ordered from small to big
    dimension. Note that inner-products necessarily requires $O(MN)$ work and
    TACO's outer-products format is dense. Finch can use a sparse outer products
    format and thus has an asymptotic advantage that becomes evident as the
    output dimensions grow. On right, we use only gustavson's algorithm and
    compare on larger matrices.}
    \label{fig:spgemm}
    \vspace{-12pt}
\end{figure}

@@

We used Finch to implement two fundamental graph algorithms: Breadth-first search (BFS) and Bellman-Ford single-source shortest path. Our BFS implementation and graphs datasets are taken from Yang et al.~\cite{yang_implementing_2018}, including road networks, characterized by bounded node degrees and long diameters, and scale-free graphs, where node degree distribution follows a power-law distribution and diameters are short.

@@

Figure~\ref{fig:graph_result} compares performance to Graphs.jl, an open-source Julia library, and the LAGraph Library, which implements graph algorithms in the language of linear algebra using GraphBLAS~\cite{mattson2019lagraph}. For the BFS algorithm, direction-optimization notably enhances performance for scale-free graphs. Although GraphBLAS uses hardwired optimizations, Finch is competitive on these benchmarks. On Bellman-Ford (with path lengths and shortest-path tree), Finch's support for multiple outputs, sparse inputs, and masks leads to superior performance over GraphBLAS (average speedup of 3.53). Note that we did not include GAP-road as it took too long to run. %speedup is arith-mean

### 2024-04-03 sections/related.tex
Update on Overleaf.

Many libraries support BLAS plus a few sparse array types, typically CSR, CSC, BCSR, Banded, and COO.
%
Examples include Scipy~\cite{virtanen2020scipy}, PETSc~\cite{abhyankarpetsc}, Armadillo~\cite{Rumengan2021}, OSKI~\cite{vuduc2005oski}, Cyclops~\cite{solomonik2013cyclops}, MKL~\cite{noauthor_developer_2024}, and Eigen~\cite{eigenweb}.
%
There are even libraries for very specific kernels and format combinations, such as SPLATT~\cite{smith2015splatt} (MTTKRP on CSF).
%
Several of these libraries also feature some graph or mesh algorithms built on sparse matrices.
%
The GraphBLAS~\cite{kepner2016mathematical} supports primitive semiring operations (operations beyond $(+, *)$, such as $(min, +)$ multiplication) which can be composed to enable graph algorithms, some of which are collected in LAGraph~\cite{mattson2019lagraph}.
%
Similarly, the Mapreduce and Hadoop platforms support operations on indexed collections~\cite{dean_mapreduce_2008}, and have been used to support graph
algorithms in the GBASE library\cite{kang2011gbase}.
%
In ML, several frameworks offer this type of support, most notably TorchSparse, which supports a few arrays and a few operations on them~\cite{tang2022torchsparse, tang2023torchsparse++}.
%
\paragraph{Compilers for Dense Data:}
%% Languages: Halide, Lift, TVM, OptiML
Outside of general purpose compilers, many compilers have been developed for optimizing dense data on a variety of control flow.
%
Perhaps the most well known example is Halide~\cite{ragan-kelley_halide_2013} and its various descendent such as TVM~\cite{chen2018tvm}, Exo~\cite{ikarashi2022exocompilation}, Elevate~\cite{hagedorn2020achieving}, and ATL~\cite{liu_verified_2022}.
%
These languages typically support most control flow except for an early break though some don't support arbitrary reading/writing or even indirect accesses.
%
Several polyhedral languages, such as Polly~\cite{grosser2012polly}, Tiramisu\cite{baghdadi2019tiramisu}, CHiLL\cite{chen2008framework}, Pluto\cite{bondhugula2008pluto}, and AlphaZ~\cite{yuki2012alphaz} offer similar capabilities in terms of control flow though they often support more irregular regions that the polyhedral framework supports.
%
These are based on ISL~\cite{verdoolaege2010isl}.
%
The density of this research represents the density of support for dense computation.

\paragraph{ Compilers for Structured Data:}
Several compilers exist for several types of structured data, often featuring separate languages for the storage of the structured data and the computation.
%
The TACO compiler originally supported just plain einsum computations\cite{kjolstad_tensor_2017}, but has been extended several times to support (single dimensional) local tensors \cite{kjolstad_tensor_2019}, breaks via semi-rings~\cite{henry_compilation_2021}, windowing and tiling \cite{senanayake2020sparse}, and convolution~\cite{won2023unified}, all as separate extensions.
%
Similarly, TACO originally support just dense and CSF like N dimensional structures, but was extended independently to support COO like structures~\cite{chou2018format}, and tree like structures~\cite{chou2022compilation}, as separate extensions.
%
The SDQL language offers a similar level of control flow~\cite{shaikhha2022functional}, but only on sparse hash tables.
%
Similarly, SDQL has been extended with a system that allows one to specify formats as queries on a set of base storage types~\cite{schleich2023optimizing} and separately by another system that describes static symmetries and other structures as predicates~\cite{ghorbani2023compiling}.
%
%Taichi is another language with a separate program and data structure specification, but
The Taichi language focuses on a single sparse data structure made from dense blocks, bit-masks, and pointers~\cite{hu_taichi_2019}.
%
%Outside of these works, the window of comparison fogs up.
%
The sparse polyhedral framework builds on CHiLL for the purpose of generating inspector/executor optimizations~\cite{strout2018sparse} though the branch of this work that specifies sparse formats separately from the computation (otherwise they are inlined into the computation manually) seems to apply mainly to einsums~\cite{zhao2022polyhedral}.
%
Second to last, SQL's classical physical/logical distinction is the classic program/format distinction, and SQL supports a huge variety of control flow constructs~\cite{kotlyar1997relational, date1989guide}.
%
However, many SQL or dataframe systems rely on BTrees, columnar, or hash tables, with only a few systems, such as Vectorwise\cite{boncz2012vectorwise}, LaraDB~\cite{hutchison2017laradb}, GMAP~\cite{tsatalos1996gmap}, or SciDB~\cite{stonebraker2013scidb} building physical layouts with other constructs based in array programming.
%
However, array based databases are a new focus given the rise of mixed ML/DB pipelines~\cite{baumann2021array,luo2018scalable}.
%
Lastly, SPIRAL focuses on recursively defined datastructures and recursively define linear algebra, and can therefore express a structure and computation that none of the systems mentioned above can: a Cooley–Tukey FFT \cite{franchetti2018spiral,franchetti2009operator}.
%
%RECUMA AND UNISPARSE are in OOPSLA R1, we should probably address these

@@

Sparse compilers have been extended to many architectures. An extension of TACO supports GPU \cite{senanayake2020sparse}, Cyclops\cite{solomonik_cyclops_2013,solomonik_sparse_2015} and SPDistal\cite{yadav_spdistal_2022-1} support distributed memory, and the Sparse Abstract Machine\cite{hsu_sparse_2022} supports custom hardware.

>>>>

Many libraries support BLAS plus a few sparse array types, typically CSR, CSC, BCSR, Banded, and COO.
%
Examples include Scipy~\cite{virtanen2020scipy}, PETSc~\cite{abhyankarpetsc}, Armadillo~\cite{Rumengan2021}, OSKI~\cite{vuduc2005oski}, Cyclops~\cite{solomonik2013cyclops}, MKL~\cite{noauthor_developer_2024}, and Eigen~\cite{eigenweb}.
%
There are even libraries for very specific kernels and format combinations, such as SPLATT~\cite{smith2015splatt} (MTTKRP on CSF).
%
Several of these libraries also feature some graph or mesh algorithms built on sparse matrices.
%
The GraphBLAS~\cite{kepner2016mathematical} supports primitive semiring operations (operations beyond $(+, *)$, such as $(min, +)$ multiplication) which can be composed to enable graph algorithms, some of which are collected in LAGraph~\cite{mattson2019lagraph}.
%
Similarly, the Mapreduce and Hadoop platforms support operations on indexed collections~\cite{dean_mapreduce_2008}, and have been used to support graph
algorithms in the GBASE library\cite{kang2011gbase}.
%
In ML, several frameworks offer this type of support, most notably TorchSparse, which supports a few arrays and a few operations on them~\cite{tang2022torchsparse, tang2023torchsparse++}.
%
\paragraph{Compilers for Dense Data:}
%% Languages: Halide, Lift, TVM, OptiML
Outside of general purpose compilers, many compilers have been developed for optimizing dense data on a variety of control flow.
%
Perhaps the most well known example is Halide~\cite{ragan-kelley_halide_2013} and its various descendent such as TVM~\cite{chen2018tvm}, Exo~\cite{ikarashi2022exocompilation}, Elevate~\cite{hagedorn2020achieving}, and ATL~\cite{liu_verified_2022}.
%
These languages typically support most control flow except for an early break though some don't support arbitrary reading/writing or even indirect accesses.
%
Several polyhedral languages, such as Polly~\cite{grosser2012polly}, Tiramisu~\cite{baghdadi2019tiramisu}, CHiLL~\cite{chen2008framework}, Pluto~\cite{bondhugula2008pluto}, and AlphaZ~\cite{yuki2012alphaz} offer similar capabilities in terms of control flow though they often support more irregular regions that the polyhedral framework supports.
%
These are based on ISL~\cite{verdoolaege2010isl}.
%
The density of this research represents the density of support for dense computation.

\paragraph{ Compilers for Structured Data:}
Several compilers exist for several types of structured data, often featuring separate languages for the storage of the structured data and the computation.
%
The TACO compiler originally supported just plain einsum computations~\cite{kjolstad_tensor_2017}, but has been extended several times to support (single dimensional) local tensors \cite{kjolstad_tensor_2019}, breaks via semi-rings~\cite{henry_compilation_2021}, windowing and tiling \cite{senanayake2020sparse}, and convolution~\cite{won2023unified}, all as separate extensions.
%
Similarly, TACO originally support just dense and CSF like N dimensional structures, but was extended independently to support COO like structures~\cite{chou2018format}, and tree like structures~\cite{chou2022compilation}, as separate extensions.
%
The SDQL language offers a similar level of control flow~\cite{shaikhha2022functional}, but only on sparse hash tables.
%
Similarly, SDQL has been extended with a system that allows one to specify formats as queries on a set of base storage types~\cite{schleich2023optimizing} and separately by another system that describes static symmetries and other structures as predicates~\cite{ghorbani2023compiling}.
%
%Taichi is another language with a separate program and data structure specification, but
The Taichi language focuses on a single sparse data structure made from dense blocks, bit-masks, and pointers~\cite{hu_taichi_2019}.
%
%Outside of these works, the window of comparison fogs up.
%
The sparse polyhedral framework builds on CHiLL for the purpose of generating inspector/executor optimizations~\cite{strout2018sparse} though the branch of this work that specifies sparse formats separately from the computation (otherwise they are inlined into the computation manually) seems to apply mainly to einsums~\cite{zhao2022polyhedral}.
%
Second to last, SQL's classical physical/logical distinction is the classic program/format distinction, and SQL supports a huge variety of control flow constructs~\cite{kotlyar1997relational, date1989guide}.
%
However, many SQL or dataframe systems rely on BTrees, columnar, or hash tables, with only a few systems, such as Vectorwise~\cite{boncz2012vectorwise}, LaraDB~\cite{hutchison2017laradb}, GMAP~\cite{tsatalos1996gmap}, or SciDB~\cite{stonebraker2013scidb} building physical layouts with other constructs based in array programming.
%
However, array based databases are a new focus given the rise of mixed ML/DB pipelines~\cite{baumann2021array,luo2018scalable}.
%
Lastly, SPIRAL focuses on recursively defined datastructures and recursively define linear algebra, and can therefore express a structure and computation that none of the systems mentioned above can: a Cooley–Tukey FFT ~\cite{franchetti2018spiral,franchetti2009operator}.
%
%RECUMA AND UNISPARSE are in OOPSLA R1, we should probably address these

@@

Sparse compilers have been extended to many architectures. An extension of TACO supports GPU~\cite{senanayake2020sparse}, Cyclops~\cite{solomonik_cyclops_2013,solomonik_sparse_2015} and SPDistal~\cite{yadav_spdistal_2022-1} support distributed memory, and the Sparse Abstract Machine~\cite{hsu_sparse_2022} supports custom hardware.
